---
title: "Introduction to GFisher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to GFisher}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(GFisher)
```

## Overview

The **GFisher** package provides fast and efficient methods for calculating Fisher combination p-values in genetic association testing. This package implements the generalized Fisher's combination test (GFisher) and its omnibus version (oGFisher), which are particularly useful for combining p-values from multiple hypothesis tests while accounting for the correlation structure among test statistics.

### Key Features

- **Flexible weighting**: Supports arbitrary non-negative weights for combining p-values
- **Generalized degrees of freedom**: Each p-value can be transformed with different degrees of freedom
- **Accounts for dependence**: Handles correlation among input test statistics through correlation matrices
- **Multiple methods**: Hybrid (HYB), Moment Ratio (MR), and Generalized Brown (GB) methods
- **Omnibus testing**: Combines multiple GFisher tests to adapt to unknown signal patterns
- **Optimized implementations**: Both R and C++ versions with substantial speedups

## Basic GFisher Test

The GFisher test generalizes Fisher's combination method by allowing flexible degrees of freedom and weights. The test statistic is:

$$S = \sum_{i=1}^n w_i F^{-1}_{d_i}(1-P_i)$$

where $F^{-1}_{d_i}$ is the inverse CDF of the chi-square distribution with $d_i$ degrees of freedom.

### Important Note on Weight Normalization

**All functions in this package automatically normalize weights to sum to 1** (i.e., $w \leftarrow w/\sum(w)$). This means:

- Providing `w = c(1, 2, 3)` is equivalent to `w = c(1/6, 2/6, 3/6)`
- This normalization improves numerical stability
- It is statistically equivalent to other normalizations like $\sum w_i = n$ in terms of p-value calculation
- The package focuses on hypothesis testing where relative weights matter, not absolute scale

This convention is consistent across all functions: `stat.GFisher()`, `p.GFisher()`, and all `*_ind` functions.

### Example 1: Standard Fisher's Method

The classical Fisher's combination test uses $d_i = 2$ and equal weights:

```{r example1}
# Simulate some p-values
set.seed(123)
n <- 10
pvals <- runif(n)  # Random p-values under null hypothesis
pvals[1:3] <- c(0.001, 0.005, 0.01)  # Add some / potential signal

# Calculate GFisher statistic (df=2 is standard Fisher's method)
stat <- stat.GFisher(pvals, df = 2, w = 1)
cat("GFisher statistic:", stat, "\n")

# For independent p-values, use fast calculation
p_value <- p.GFisher_ind_w1(stat, df = rep(2, n))
cat("P-value (under independency):", p_value, "\n")
```

### Example 2: Varying Degrees of Freedom

GFisher allows different degrees of freedom for each p-value transformation.
(Examples 2 and 3 use `p.GFisher_ind()`, whose exact method needs the optional
`coga` package; they are skipped automatically where `coga` is unavailable.)

```{r example2, eval = requireNamespace("coga", quietly = TRUE)}
# Use different df values
df_values <- rep(c(1, 2, 3, 4), length.out = n)

# Calculate statistic with varying df
stat2 <- stat.GFisher(pvals, df = df_values, w = 1)
cat("GFisher statistic (varying df):", stat2, "\n")

# P-value for independent case
p_value2 <- p.GFisher_ind(stat2, df = df_values, w = rep(1, n))
cat("P-value (independent, varying df):", p_value2, "\n")
```

### Example 3: Weighted Combination

Different weights can emphasize certain p-values:

```{r example3, eval = requireNamespace("coga", quietly = TRUE)}
# Assign higher weights to first few p-values
weights <- c(rep(3, 3), rep(1, n - 3))

# Calculate weighted statistic
stat3 <- stat.GFisher(pvals, df = 2, w = weights)
cat("Weighted GFisher statistic:", stat3, "\n")

# P-value with weights
p_value3 <- p.GFisher_ind(stat3, df = rep(2, n), w = weights)
cat("P-value (weighted):", p_value3, "\n")
```

## Accounting for Dependence

In genetic association studies, test statistics are often correlated due to linkage disequilibrium (LD). GFisher accounts for this through correlation matrices.

### Example 4: Correlated Test Statistics

```{r example4}
# Create correlation matrix (e.g., from LD structure)
M <- matrix(0.3, n, n) + diag(0.7, n, n)

# Generate correlated z-scores
set.seed(456)
M_chol <- chol(M)
z_scores <- rnorm(n) %*% M_chol
pvals_cor <- 2 * pnorm(-abs(z_scores))  # Two-sided p-values

# Calculate statistic
stat_cor <- stat.GFisher(pvals_cor, df = 2, w = 1)

# Calculate p-value accounting for dependence (Hybrid method)
p_hyb <- p.GFisher(stat_cor, df = 2, w = 1, M = M, method = "HYB")
cat("P-value (HYB, dependent):", p_hyb, "\n")

# Compare with independence assumption (would be incorrect!)
p_ind <- p.GFisher_ind_w1(stat_cor, df = rep(2, n))
cat("P-value (independent assumption - WRONG!):", p_ind, "\n")
cat("Ratio (dependent/independent):", p_hyb / p_ind, "\n")
```

The p-value accounting for dependence is typically larger (more conservative) than incorrectly assuming independence.

## Comparison of Methods

GFisher supports three methods for p-value calculation:

- **HYB (Hybrid)**: Fast quadratic approximation using first four moments (two-sided p-values only)
- **MR (Moment Ratio)**: Simulation-based with moment matching (more accurate for extreme tails)
- **GB (Generalized Brown)**: Brown's method with calculated variance (simplest approximation)

### Example 5: Method Comparison

```{r example5}
# Test statistic
stat_test <- stat.GFisher(pvals_cor, df = 2, w = 1)

# Compare methods
p_hyb <- p.GFisher(stat_test, df = 2, w = 1, M = M, method = "HYB")
p_mr <- p.GFisher(stat_test, df = 2, w = 1, M = M, method = "MR", nsim = 1e4)
p_gb <- p.GFisher(stat_test, df = 2, w = 1, M = M, method = "GB")

cat("Method comparison:\n")
cat("  HYB:", p_hyb, "\n")
cat("  MR: ", p_mr, "\n")
cat("  GB: ", p_gb, "\n")
```

**Recommendation**: Use HYB for speed and accuracy with two-sided p-values. Use MR for extreme tail probabilities or one-sided p-values.

## Omnibus GFisher Test (oGFisher)

The omnibus test combines multiple GFisher statistics with different configurations (degrees of freedom and weights) to adapt to unknown signal patterns. This is particularly powerful when the optimal test configuration is unknown.

### Example 6: oGFisher with Multiple Configurations

```{r example6}
# Define multiple test configurations
# Each row represents a different GFisher test
nGF <- 5  # Number of GFisher tests
DF <- matrix(c(
  rep(1, n),  # Test 1: df=1
  rep(2, n),  # Test 2: df=2 (Fisher's method)
  rep(3, n),  # Test 3: df=3
  rep(4, n),  # Test 4: df=4
  c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5)  # Test 5: varying df
), nrow = nGF, byrow = TRUE)

W <- matrix(rep(1, nGF * n), nrow = nGF)  # Equal weights

# Calculate oGFisher statistics
result <- stat.oGFisher(pvals_cor, DF, W, M, method = "HYB")

cat("Individual GFisher statistics:\n")
print(result$STAT)

cat("\nIndividual p-values:\n")
print(result$PVAL)

cat("\nMinimum p-value:", result$minp, "\n")
cat("CCT statistic:", result$cct, "\n")
```

### Example 7: oGFisher P-Value

```{r example7}
# Calculate omnibus p-value using Cauchy combination
result_cct <- pval.oGFisher(pvals_cor, DF, W, M, method = "HYB", combine = "cct")
cat("oGFisher p-value (CCT):", result_cct$pval, "\n")

# Alternative: minimum p-value with MVN distribution
result_mvn <- pval.oGFisher(pvals_cor, DF, W, M, method = "HYB", combine = "mvn")
cat("oGFisher p-value (MVN):", result_mvn$pval, "\n")
```

## Genetic Association Example

Let's simulate a more realistic scenario: testing association of rare variants in a gene.

### Example 8: Gene-Based Rare Variant Test

```{r example8}
# Simulate a gene with 15 rare variants
n_variants <- 15
set.seed(789)

# Simulate LD structure (higher correlation for nearby variants)
# Create block-diagonal structure
M_ld <- matrix(0.1, n_variants, n_variants)
diag(M_ld) <- 1

# Add blocks of higher correlation (nearby variants)
for (i in 1:3) {
  idx <- ((i-1)*5 + 1):(i*5)
  M_ld[idx, idx] <- 0.6
  diag(M_ld[idx, idx]) <- 1
}

# Simulate z-scores with this LD structure
z_variants <- rnorm(n_variants) %*% chol(M_ld)

# Add signal to some variants (causal variants)
z_variants[c(2, 3, 8)] <- z_variants[c(2, 3, 8)] + 4.5  # Add signal

# Convert to two-sided p-values
pvals_variants <- 2 * pnorm(-abs(z_variants))

cat("P-values for variants:\n")
print(round(pvals_variants, 5))

# Test 1: Standard Fisher's method accounting for LD
stat_fisher <- stat.GFisher(pvals_variants, df = 2, w = 1)
p_fisher <- p.GFisher(stat_fisher, df = 2, w = 1, M = M_ld, method = "HYB")
cat("\nFisher's method p-value:", p_fisher, "\n")

# Test 2: Weighted by inverse MAF (common in rare variant analysis)
# Simulate MAF (minor allele frequencies)
maf <- runif(n_variants, 0.001, 0.05)
weights_maf <- 1 / sqrt(maf * (1 - maf))  # Madsen-Browning weights
weights_maf <- weights_maf / sum(weights_maf) * n_variants  # Normalize

stat_weighted <- stat.GFisher(pvals_variants, df = 2, w = weights_maf)
p_weighted <- p.GFisher(stat_weighted, df = 2, w = weights_maf, M = M_ld, method = "HYB")
cat("Weighted test p-value:", p_weighted, "\n")

# Test 3: Omnibus test to adapt to signal
DF_gene <- rbind(
  rep(1, n_variants),
  rep(2, n_variants),
  rep(3, n_variants)
)
W_gene <- rbind(
  rep(1, n_variants),
  weights_maf,
  sqrt(weights_maf)
)

result_omnibus <- pval.oGFisher(pvals_variants, DF_gene, W_gene, M_ld,
                                 method = "HYB", combine = "cct")
cat("Omnibus test p-value:", result_omnibus$pval, "\n")

cat("\nConclusion: oGFisher adapts to the signal pattern and provides",
    "\na powerful omnibus test for genetic association.\n")
```

## Performance: R vs C++

The package includes optimized C++ implementations for computational bottlenecks. Here's a simple benchmark:

```{r benchmark, eval=FALSE}
# This is not evaluated in the vignette to save time
# Run this code manually to see speedups

library(microbenchmark)

n <- 20
M <- matrix(0.3, n, n) + diag(0.7, n, n)
pvals <- runif(n)
q <- stat.GFisher(pvals, df = 2, w = 1)

# Compare R vs C++ for MR method
microbenchmark(
  R_version = p.GFisher(q, df = 2, w = 1, M = M, method = "MR",
                        nsim = 1e5, use.cpp = FALSE),
  Cpp_version = p.GFisher(q, df = 2, w = 1, M = M, method = "MR",
                          nsim = 1e5, use.cpp = TRUE),
  times = 10
)

```

See the `benchmarks/` directory in the package source for comprehensive benchmark scripts.

## Summary

The GFisher package provides:

1. **Flexible combination tests**: Generalized Fisher's method with varying df and weights
2. **Dependence handling**: Accounts for correlation structure among test statistics
3. **Omnibus testing**: Adapts to unknown signal patterns
4. **Multiple methods**: HYB, MR, and GB for different scenarios
5. **High performance**: C++ implementations for speed-critical operations
6. **Special cases**: Optimized functions for independent p-values

For genetic association testing, GFisher is particularly useful for:

- Gene-based rare variant tests
- Combining results from multiple variants in LD
- Adaptive tests when optimal weights are unknown
- Meta-analysis with correlated test statistics

## References

Zhang, H. and Wu, Z. (2022). The generalized Fisher's combination and accurate p-value calculation under dependence. *Biometrics*, 79(2), 1159-1172. doi:10.1111/biom.13634

---
